Activity Monitoring
===========================

### Loading the data and preparing for analysis
I load and run a summary of the activity data. 

```{r load_data, echo=TRUE}

#Load libraries and data
library(dplyr)
data <- read.csv("activity.csv")

#Instead of processing the data, running a summnary
summary(data)

```

### What is mean total number of steps taken per day?
Ignoring missing values in this part of the analysis

```{r mean_total_steps, echo =TRUE}

#Calculate the total number of steps taken per day
steps_per_day <- data%>% group_by(date) %>%
  summarise(steps = sum(steps, na.rm = TRUE))

#Make a histogram of the total number of steps taken each day
Steps <- steps_per_day$steps
hist(Steps)

#Calculate and report the mean and median of the total number of steps taken per day
mean_steps <- mean(Steps, na.rm=TRUE)
median_steps <- median(Steps, na.rm=TRUE)

```
The mean and median of the total number of steps taken per day are `r mean_steps` and `r median_steps`, respectively.



### What is the average daily activity pattern?

```{r avg_daily_pattern, echo=TRUE}

#Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
steps_per_interval <- data%>% group_by(interval) %>%
  summarise(mean_steps = mean(steps, na.rm=TRUE))

Interval <- steps_per_interval$interval
Steps <- steps_per_interval$mean_steps

plot(Interval, Steps,
     main ="5-minute intervals and the average number of steps taken",
     type="l",
     col ="blue")

```

### Imputing missing values
Accounting for the presence of missing days, which may introduce bias into some calculations or summaries of the data.

```{r missing_values, echo=TRUE}

#Calculate and report the total number of missing values in the dataset
NAs <- data %>% summarise(NAs = sum(is.na(steps)))

#Create a new dataset with the missing data filled in with the mean for that day 5-minute interval
data_filled <- data %>% left_join(steps_per_interval, by = "interval") %>%
  mutate(adj_steps = ifelse(is.na(steps), mean_steps, steps))

#Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Reusing code from above but changing data to data_filled and steps to adjusted steps.

#Calculate the total number of steps taken per day
adj_steps_per_day <- data_filled%>% group_by(date) %>%
  summarise(adj_steps = sum(adj_steps, na.rm = TRUE))

#Make a histogram of the total number of steps taken each day
Adjusted_Steps <- adj_steps_per_day$adj_steps
hist(Adjusted_Steps)

#Calculate and report the mean and median of the total number of steps taken per day
adj_mean_steps <- mean(Adjusted_Steps, na.rm=TRUE)
adj_median_steps <- median(Adjusted_Steps, na.rm=TRUE)

# Calculate the differences between the adjusted mean and median and the originals
mean_diff <- adj_mean_steps - mean_steps
median_diff <- adj_median_steps - median_steps

```

There are `r NAs` missing values in the dataset. Imputing the missing values increase the mean and median by `r mean_diff` and `r median_diff`, respectively. You can compare the two histograms and see that the adjusted data cluster more in the center of the histogram.



### Are there differences in activity patterns between weekdays and weekends?
Using the dataset with the filled-in missing values for this part.

```{r weekends, echo=TRUE}
#Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
data_filled <- data_filled%>% mutate(
  day_type = ifelse(weekdays(as.Date(date), abbreviate = FALSE) == "Saturday" | 
                      weekdays(as.Date(date), abbreviate = FALSE) == "Sunday", "Weekend", "Weekday")
)

#Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data. Copying code from above but changing to adjusted/filled.

#Weekday
filled_steps_per_interval_weekday <- data_filled%>% filter(day_type == "Weekday") %>%
  group_by(interval) %>%
  summarise(filled_mean_steps = mean(adj_steps, na.rm=TRUE))

Filled_Interval_Weekday <- filled_steps_per_interval_weekday$interval
Filled_Steps_Weekday <- filled_steps_per_interval_weekday$filled_mean_steps

#Weekend
filled_steps_per_interval_weekend <- data_filled%>% filter(day_type == "Weekend") %>%
  group_by(interval) %>%
  summarise(filled_mean_steps = mean(adj_steps, na.rm=TRUE))

Filled_Interval_Weekend <- filled_steps_per_interval_weekend$interval
Filled_Steps_Weekend <- filled_steps_per_interval_weekend$filled_mean_steps

#Plot
par(mfrow=c(2,1))
plot(Filled_Interval_Weekday, Filled_Steps_Weekday,
     main ="weekday",
     type="l",
     col ="blue")
plot(Filled_Interval_Weekend, Filled_Steps_Weekend,
     main ="weekend",
     type="l",
     col ="blue")



```

